For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
stuff

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,A\ra Con^D\,(\mathsf{ext}\,\GG\,A) \\
  & U^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
  & El^D && : (\GG^D:Con^D\,\GG)\ra\,Ty^D\,({ext}^D\,\GG^D\,(U^D\,\GG^D))\,(\mathsf{El}\,\GG)
%  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
\end{alignat*} xxx
Above we used implicit quantifications for $\GG:\Con$ and
$A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\GG$ before its explicit parameter of type
$Con^D\,\GG$.

@@

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
extensional type theory however this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle is that
$\ll\blank\rr$ for types refers to $\ll\blank\rr$ for contexts, hence
this function is recursive-recursive.

@@

In this paper we show that any model of extensional type theory with
indexed W-types supports the theory of IIT signatures, hence all
IITs. The difficulty is that the theory of IIT signatures is a QIIT,
it is both inductive-inductive and has equality constructors. However
it can be seen as the well-typed syntax of a small type theory without
any computation rules. Hence we can represent the syntax of normal
forms without quotienting. We construct this well-typed normal syntax
using preterms and typing relations from indexed W-types. Finally, we
prove the elimination principle in the style of the initiality proof
of Streicher \cite{streichersemantics}.

Just as ibid., we only consider finitary IITs, that is, constructors
can only have a finite number of recursive arguments. An example
constructor for $\Con$--$\Ty$ which is not allowed is the following:
\[
  \Pi_\infty : (\GG : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\GG) \rightarrow \Ty\,\GG
\]

TODO: list of contributions here. Notion of signature free from encoding.

@@

\subsection{Structure of the Paper}

We first define a theory of IIT signatures in which every context is a signature
(Section \ref{sec:theory_of_signatures}). We present this as an algebraic
\cite{ttintt} theory, specifically as a quotient inductive-inductive type
(QIIT). In Section \ref{sec:ambroise} we show that this QIIT can be constructed
using inductive types, UIP and function extensionality. In Section
\ref{sec:constructingiits} we describe the construction of IITs from the syntax
of IIT signatures.

\subsection{Related Work}

The current work builds heavily on the work of Kaposi et al.\
\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
inductive-inductive types (QIITs); we reuse both QIIT syntax and
semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well. We also make use of the extesion to
infinitary QIITs \cite{large_inf_qiit} to derive the specification of
the elimination principle for the theory of IIT signatures. TODO: also
for specifying signatures.

@@

@@

\subsection{Notation and Formalisation}

\begin{definition}[Model of extensional type theory (ETT)]
  By a model of ETT we mean a category with families (CwF)
  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
  predicative hierarchy of universes closed under the following type
  formers: $\Pi$, $\Sigma$, $\bot$, $\top$, $\Bool$, indexed W-types
  \cite{indexedcont} and an identity type with uniqueness of identity
  proofs and equality reflection. TODO: remove indexed W-types, bot
  and bool.
\end{definition}

We will use Agda-like type theoretic syntax to work in the internal
language of models of ETT:
\begin{itemize}
\item Universes are written $\Set_i$. We usually omit level indices in this paper.
\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
  non-dependent. We sometimes omit function arguments, by implicitly
  generalising over variables.
\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
  $\sum\limits_{x} B$ when we want to leave the type of the first
  projection implicit. Projections are either named or given by
  $\proj_1$ and $\proj_2$. We use $A \times B$ for non-dependent
  pairs.
\item We will define indexed W-types \cite{indexedcont} by listing
  their constructors and using pattern matching to eliminate from
  them.
\item The equality (identity) type is written $t = u$, it has a
  constructor $\refl : t = t$, and equality reflection, hence we use
  the same $=$ sign for definitional equality. We occasionally
  indicate by $_{e_1,\dots,e_n \#}t$ that $t$ is well-typed thanks to the
  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
  function extensionality, expressing
  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
\end{itemize}

The contents of Sections \ref{sec:theory_of_signatures} and
\ref{sec:ambroise} were formalised in Agda, the formalisation is
available at \url{https://github.com/amblafont/UniversalII}. Agda's
pattern matching mechanism implies uniqueness of identity proofs, we
assumed function extensionality as an axiom and used rewrite rules
\cite{cockxsprinkles} to obtain limited equality reflection.

@@

\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
  A model of ETT supports IITs if for any signature $(sig,sigc):\Sign$
  there is a
  \[
    \con_{sig}: ({sig}\,\ADS)^\A
  \]
  and an
  \[
    \elim_{sig} : (\gamma^D:({sig}\,\ADS)^\D\,\con_{sig})\ra ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{definition}
In other words, for any signature, we have an algebra called $\con$
(constructors) and for any displayed algebra over the constructors, we
have a section (called the eliminator).

One can check that \Cref{sec:hasiits} gives the right

@@

\section{Further Work}

>>>>

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,A\ra Con^D\,(\mathsf{ext}\,\GG\,A) \\
  & U^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
  & El^D && : (\GG^D:Con^D\,\GG)\ra\,Ty^D\,({ext}^D\,\GG^D\,(U^D\,\GG^D))\,(\mathsf{El}\,\GG)
%  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
\end{alignat*}
Above we used implicit quantifications for $\GG:\Con$ and
$A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\GG$ before its explicit parameter of type
$Con^D\,\GG$.

@@

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
extensional type theory however this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.

@@

In this paper we show that any model of extensional type theory with
indexed W-types supports the theory of IIT signatures, and as a
consequence all IITs. The difficulty in this construction is that the
theory of IIT signatures is itself a QIIT, it is both
inductive-inductive and has equality constructors. However it can be
seen as the well-typed syntax of a small type theory without any
computation rules. Hence we can represent the syntax of normal forms
without quotienting. We construct this well-typed normal syntax using
preterms and typing relations from indexed W-types. Finally, we prove
the elimination principle in the style of the initiality proof of
Streicher \cite{streichersemantics}.

Just as \cite{Kaposi:2019:CQI:3302515.3290315}, we only consider
finitary IITs, that is, constructors can only have a finite number of
recursive arguments. An example constructor for $\Con$--$\Ty$ which is
not allowed is the following:
\[
  \Pi_\infty : (\GG : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\GG) \rightarrow \Ty\,\GG
\]

\paragraph*{Structure of paper and list of contributions}

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. We show that if a model of ETT
  supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs in the above
  sense: \Cref{th:tosToIITs}. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{TODO}), and hence, all IITs (\Cref{TODO}).
\end{itemize}
We conclude in \Cref{sec:conclusions}.

@@

\subsection{Related Work}
\label{sec:related}

The current work builds heavily on the work of Kaposi et al.\
\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
inductive-inductive types (QIITs); we reuse both QIIT syntax and
semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well. We also make use of the extesion to
infinitary QIITs \cite{large_inf_qiit} to derive the specification of
the elimination principle for the theory of IIT signatures.

@@

Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18} construct
inductive types using a restricted Church encoding in a type theory
with an impredicative universe. We use the predicative version of
their encoding to define IIT signatures.

@@

\subsection{Notation and Formalisation}
\label{sec:notation}

\begin{definition}[Model of extensional type theory (ETT)]\label{def:ett}
  By a model of ETT we mean a category with families (CwF)
  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
  predicative hierarchy of universes closed under the following type
  formers: $\Pi$, $\Sigma$, $\top$ and an identity type with
  uniqueness of identity proofs and equality reflection.
\end{definition}

We will use Agda-like type theoretic syntax to work in the internal
language of models of ETT:
\begin{itemize}
\item Universes are written $\Set_i$. We usually omit level indices in this paper.
\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
  non-dependent. We sometimes omit function arguments, by implicitly
  generalising over variables.
\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
  $\sum\limits_{x} B$ when we want to leave the type of the first
  projection implicit. Projections are either named or given by
  $\proj_1$ and $\proj_2$. We use $A \times B$ for non-dependent
  pairs.
\item The unit type $\top$ has the constructor $\tt$ which is
  definitionally equal to all elements of $\top$.
\item The equality (identity) type is written $t = u$, it has a
  constructor $\refl : t = t$, and equality reflection, hence we use
  the same $=$ sign for definitional equality. We occasionally
  indicate by $_{e_1,\dots,e_n \#}t$ that $t$ is well-typed thanks to
  the equalities $e_1$,\dots,$e_n$. To construct proofs, sometimes we
  write equational reasoning, e.g. $f a \overset{e}{=} f b$ where
  $e : a = b$. We also have uniqueness of identity proofs (UIP),
  expressing $(e : t = t)\ra e = \refl$. Note that function
  extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$ is
  derivable.
\end{itemize}

The contents of Section \ref{sec:ambroise} were formalised in Agda,
the formalisation is available at
\url{https://github.com/amblafont/UniversalII}. Agda's pattern
matching mechanism implies uniqueness of identity proofs, we assumed
function extensionality as an axiom and used rewrite rules
\cite{cockxsprinkles} to obtain limited equality reflection.

@@

\begin{definition}[A model of ETT supports IITs]\label{def:hasiits}
  A model of ETT supports IITs if for any signature $(sig,sigc):\Sign$
  there is a
  \[
    \con_{sig}: ({sig}\,\ADS)^\A
  \]
  and an
  \[
    \elim_{sig} : (\gamma^D:({sig}\,\ADS)^\D\,\con_{sig})\ra ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{definition}
In other words, for any signature, we have an algebra called $\con$
(constructors) and for any displayed algebra over the constructors, we
have a section (called the eliminator).

One can check that \Cref{def:hasiits} gives the right

@@

\section{Conclusions}
\label{sec:conclusions}

TODO
